\chapter*{Publications}
As required by \S 5(3) of the 
\\

\begin{addmargin}{0.07\textwidth}
\textit{Promotionsordnung für die Fakultät für Mathematik, Informatik und Naturwissenschaften der Rheinisch-Westfälischen Technischen Hochschule Aachen vom 27.09.2010 in der Fassung der zweiten Ordnung zur Änderung der Promotionsordnung vom 30.06.2014 (veröffentlicht als Gesamtfassung)},
\end{addmargin} 
\vspace{0.5cm}

\noindent
a declaration of results that are published by the author as well as particular contributions to co-authored articles follows.

I am the main author of the following articles which are co-authored with Ivan G. Costa. 
As my adviser, Ivan G. Costa supported me in all stages of research.
\\

\begin{addmargin}{0.07\textwidth}
Manuel Allhoff, Kristin Seré, Heike Chauvistré, Qiong Lin, Martin Zenke, and Ivan G. Costa.
\textit{Detecting differential peaks in ChIP-seq signals with ODIN.}
\\

\noindent
Manuel Allhoff, Juliana F. Pires, Kristin Seré, Martin Zenke, and Ivan G. Costa.
\textit{Differential Peak Calling of ChIP-seq Signals with Replicates with THOR.}
\end{addmargin} 
\vspace{0.5cm}

\noindent
I implemented ODIN, THOR as well as the simulation algorithm.
Moreover, I performed all data analysis experiments.
All chapters of this thesis have been written by me.
Only Section~\ref{sec_expmethod_rsnps} and Section~\ref{res_ident_rsnps} (Identifying rSNPs) are equally shared by Juliana F. Pires and me.

The following article did not directly contribute to this thesis, but shaped a general understanding of next-generation sequencing analysis.
\\

\begin{addmargin}{0.07\textwidth}
Manuel Allhoff, Alexander Sch\"onhuth, Marcel Martin, Ivan G. Costa, Sven Rahmann, and Tobias Marschall. 
\textit{Discovering motifs that induce sequencing errors.}
\end{addmargin} 
