\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[body={6.25in,8.85in}]{geometry}
\usepackage{url}
\usepackage[scaled]{helvet}
\usepackage[authoryear,round,longnamesfirst]{natbib}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage[nolists,nomarkers]{endfloat}

\begin{document}
\title{Studentische Hilfskraft: Next-Generation High-throughput Genomic Signal Processing} \date{}

\author{Eduardo G. Gusmao and Ivan G. Costa\\
IZKF Computational Biology Research Group -- Institute for Biomedical Engineering
}

\thispagestyle{empty}

\maketitle

\setlength{\parskip}{0.5cm}

\section{Unser Profil}
\label{sec:unserprofil}

The main focus of the IZKF Computational Biology Research Group is on the development of strategies to extract meaningful information from high-throughput genomic data. Improvements on next-generation sequencing technologies have enabled the production of massive amounts of genomic data. To analize these datasets it is necessary to create robust mathematical and efficient computational approaches. Our group is composed of a interdisciplinary team with computer scientists, bioinformaticians and biologists. 

\section{Ihr Profil}
\label{sec:ihrprofil}

\begin{itemize}
  \item Currently enrolled in on of the following areas: electronic engineering, computer science, mathematics or physics.
  \item Very good knowledge of signal processing and time-series data analysis.
  \item Very good knowledge of at least one programming language, preferably: python, R, matlab or C.
  \item Sehr gute Englischkenntnisse.
  \item Sehr hohe Motivation und Einsatzbereitschaft.
  \item Ausgepr{\"a}gte kommunikative F{\"a}higkeiten und Interesse an Teamarbeit.
\end{itemize}

\section{Ihre Aufgaben}
\label{sec:ihreaufgaben}

Currently, we have a method based on  hidden Markov models that is able to identify particular patterns on multivariate genomic signals. The goals of this research projects are:
\begin{enumerate}
  \item Compare the patterns identified by our method with patterns obtained by applying signal processing techniques.
  \item Use signal processing skills to filter the signal in order to allow more accurate identification of interesting patterns.
\end{enumerate}

\section{Unser Angebot}
\label{sec:unserangebot}

\begin{itemize}
  \item Mitarbeit in einem spannenden und "multidisciplinar" Forschungsfeld.
  \item Praktische Arbeitserfahrungen neben dem Studium.
  \item Flexible Arbeitszeiten.
%  \item The possibility of publishing all discoveries in high-impact journals.
  \item Die Stelle ist zum n{\"a}chstm{\"o}glichen Zeitpunkt zu besetzen und befristet auf zun{\"a}chst 6 Monate (with possibility of extension).
  \item Die regelm{\"a}{\ss}ige Wochenarbeitszeit betr{\"a}gt 10 Stunden.
  \item Die Eingruppierung richtet sich nach der Richtlinie f{\"u}r studentische und wissenschaftliche Hilfskr{\"a}fte.

% XXX - IC - I think these are already standar in RWHT. One dont need to put extra. 
%  \item Die RWTH ist als familiengerechte Hochschule zertifiziert. 
%  \item Wir wollen an der RWTH Aachen besonders die Karrieren von Frauen f{\"o}rdern und freuen uns daher {\"u}ber Bewerberinnen.
%  \item Frauen werden bei gleicher Eignung, Bef{\"a}higung und fachlicher Leistung bevorzugt ber{\"u}cksichtigt, sofern sie in der Organisationseinheit unterrepr{\"a}sentiert sind und sofern nicht in der Person eines Mitbewerbers liegende Gr{\"u}nde {\"u}berwiegen. 
%  \item Bewerbungen geeigneter schwerbehinderter Menschen sind ausdr{\"u}cklich erw{\"u}nscht.
\end{itemize}

\section{Ihr/e Ansprechpartner/in}
\label{sec:ansprechpartner}

F{\"u}r Vorabinformationen steht Ihnen \\
Herr Eduardo Gusmao \\
Tel.: +49 (0) 241 80-80271 \\
E-Mail: eduardo.gusmao@rwth-aachen.de

\noindent
zur Verf{\"u}gung. \\
Nutzen Sie auch unsere Webseiten zur Information: \\
http://costalab.org/ \\

\noindent
Ihre Bewerbung richten Sie bitte bis zum 31.01.2014 an \\
Dr. Ivan Costa / Eduardo Gusmao \\
Helmholtz Institute for Biomedical Engineering \\
Centre of Medical Technology (MTZ) \\
Pauwelsstr. 19 / Room 3.01\\
52074 Aachen

\noindent
Gerne k{\"o}nnen Sie Ihre Bewerbung auch per E-Mail an eduardo.gusmao@rwth-aachen.de senden. Bitte beachten Sie, dass Gef{\"a}hrdungen der Vertraulichkeit und der unbefugte Zugriff Dritter bei einer Kommunikation per unverschl{\"u}sselter E-Mail nicht ausgeschlossen werden k{\"o}nnen.

\bibliographystyle{natbib}
\bibliography{document}

\end{document}


