
\chapter*{Glossary}

\markright{\MakeUppercase{\nomname}}

\begin{multicols}{2}
\begin{footnotesize}

\nomenclature{bp}{Base pair.} % OK
\nomenclature{TFBS}{Transcription factor binding site.} % OK
\nomenclature{TF}{Transcription factor.} % OK
\nomenclature{HMM}{Hidden Markov model.} % OK
\nomenclature{PFM}{Position frequency matrix.} % OK
\nomenclature{PWM}{Position weight matrix.} % OK
\nomenclature{DHS}{DNase hypersensitivity site.} % OK
\nomenclature{MPBS}{Motif-predicted binding site.} % OK
\nomenclature{NGS}{Next-generation sequencing.} % OK
\nomenclature{TC}{Tag count.} % OK
\nomenclature{FS}{Footprint score.} % OK
\nomenclature{FC}{Gene expression fold change.} % OK
\nomenclature{ROC}{Receiver operating characteristics.} % OK
\nomenclature{AUC}{Area under the ROC curve.} % OK
\nomenclature{PR}{Precision-recall.} % OK
\nomenclature{AUPR}{Area under the PR curve.} % OK
\nomenclature{FPR}{False positive rate.} % OK
\nomenclature{KS}{Kolmogorov-Smirnov.} % OK

\nomenclature{DNA}{Deoxyribonucleic acid.} % OK
\nomenclature{RNA}{Ribonucleic acid.} % OK
\nomenclature{ChIP-seq}{Chromatin immunoprecipitation followed by massive sequencing.} % OK
\nomenclature{DNase-seq}{DNase~I digestion followed by massive sequencing.} % OK
\nomenclature{ENCODE}{Encyclopedia of DNA Elements.} % OK
\nomenclature{MACS}{Model-based analysis for ChIP-Seq software.} % OK
\nomenclature{BWA}{Burrows-Wheeler aligner software.} % OK
\nomenclature{FIMO}{Find individual motif occurrences software.} % OK
\nomenclature{HINT}{HMM-based identification of TF footprints.} % OK
\nomenclature{RGT}{Regulatory genomics toolbox.} % OK
\nomenclature{PIQ}{Protein interaction quantification.} % OK
\nomenclature{FLR}{Footprint likelihood ratio.} % OK
\nomenclature{GEO}{Gene expression omnibus.} % OK
\nomenclature{SRA}{Sequence read archive.} % OK
\nomenclature{TSS}{Transcription start site.} % OK
\nomenclature{IQR}{Interquartile region.} % OK
\nomenclature{FP-Exp}{Correlation of KS test statistic and gene expression FC.} % OK
\nomenclature{OBS}{Observed \emph{vs} bias signal.} % OK
\nomenclature{DREME}{Discriminative regular expression motif elicitation.} % OK
\nomenclature{CENTRIMO}{Local motif enrichment analysis.} % OK
\nomenclature{IUPAC}{International union of pure and applied chemistry.} % OK
\nomenclature{DC}{Dendritic cell.} % OK
\nomenclature{MPP}{Multipotent progenitor cell.} % OK
\nomenclature{CDP}{Common dendritic cell progenitor.} % OK
\nomenclature{cDC}{Classical dendritic cell.} % OK
\nomenclature{pDC}{Plasmacytoid dendritic cell.} % OK
\nomenclature{SH}{Single-hit DNase-seq protocol.} % OK
\nomenclature{DH}{Double-hit DNase-seq protocol.} % OK
\nomenclature{NK}{Naked (deproteinized) DNA DNase-seq protocol.} % OK

\printnomenclature[1.8cm] % [] = distance between entry and description
\label{nom} % target name for links to glossary

\end{footnotesize}
\end{multicols}

