\chapter*{List of Publications}

\subsubsection*{First Author}

\begin{addmargin}{0.05\textwidth}
\noindent
\textbf{Gusmao EG}, Allhoff M, Zenke M, Costa IG (2016). Analysis of computational footprinting methods for DNase sequencing experiments. \textit{Nature Methods}, 13(4):303--309.\\ \\
\noindent
\textbf{Gusmao EG}, Dieterich C, Zenke M, Costa IG (2014). Detection of active transcription factor binding sites with the combination of DNase hypersensitivity and histone modifications. \textit{Bioinformatics}, 30(22):3143--3151.\\ \\
\noindent
\textbf{Gusmao EG}, de Souto MCP (2014). Issues on Sampling Negative Examples for Predicting Prokaryotic Promoters. \textit{Proceedings of the International Joint Conference on Neural Networks (IJCNN 2014)}, 494--501.
\end{addmargin} 


\subsubsection*{Collaborations}

\begin{addmargin}{0.05\textwidth}
\noindent
Kolovos P, Georgomanolis T, Nikolic M, Koeferle A, Larkin JD, Feuerborn A, van Ijcken WF, \textbf{Gusmao EG}, Costa IG, Cook PR, Grosveld FG, Papantonis A (2016). Enhancer hijacking reveals a multimodal role of NF-$\kappa$B during the immediate-early inflammatory response. \textit{(in review)}.\\ \\
\noindent
Lin Q, Chauvistr\'{e} H, Costa IG, \textbf{Gusmao EG}, Mitzka S, H\"{a}enzelmann S, Baying B, Klisch T, Moriggl R, Hennuy B, Smeets H, Hoffmann K, Benes V, Ser\'{e} K, Zenke M (2015). Epigenetic and transcriptional architecture of dendritic cell development. \textit{Nucleic Acids Research}, 43(20):9680--9693.\\ \\
\noindent
Schemionek M, Herrmann O, Reher MM, Chatain N, Schubert C, Costa IG, \linebreak H\"{a}nzelmann S, \textbf{Gusmao EG}, Kintsler S, Braunschweig T, Hamilton A, Helgason GV, Copland M, Schwab A, M\"{u}ller-Tidow C, Li S, Holyoake TL, Br\"{u}mmendorf TH, \linebreak Koschmieder S (2015). MTSS1 is a critical epigenetically regulated tumor suppressor in CML. \textit{Leukemia}, 30(4):823--832. \\ \\
\noindent
H\"{a}nzelmann S, Beier F, \textbf{Gusmao EG}, Koch CM, Hummel S, Charapitsa I, Joussen S, Benes V, Br\"{u}mmendorf TH, Reid G, Costa IG, Wagner W (2015). Replicative senescence is associated with nuclear reorganization and with DNA methylation at specific transcription factor binding sites. \textit{Clinical Epigenetics}, 7(1):19. \\ \\
\noindent
Ullius A, L\"{u}scher-Firzlaff J, Costa IG, Walsemann G, Forst AH, \textbf{Gusmao EG}, Kapelle K, Kleine H, Kremmer K, Vervoorts J, L\"{u}scher B (2014). The interaction of MYC with the trithorax protein ASH2L promotes gene transcription by regulating H3K27 modification. \textit{Nucleic Acids Research}, 42(11):6901--6920. \\ \\
\end{addmargin} 


