\chapter*{Publications}
As required by \S 5(3) of the \\

\begin{addmargin}{0.05\textwidth}
\textit{Promotionsordnung f\"ur die Fakult\"at f\"ur Mathematik, Informatik und Naturwissenschaften der Rheinisch-Westf\"alischen Technischen Hochschule Aachen vom 27.09.2010 in der \linebreak Fassung der zweiten Ordnung zur \"anderung der Promotionsordnung vom 30.06.2014 \linebreak (ver\"offentlicht als Gesamtfassung)},
\end{addmargin} 
\vspace{0.5cm}

\noindent
a declaration of results that are published by the author as well as particular contributions to co-authored publications follows. \\

I am the main author of the following publications, which are co-authored with Dr. Ivan G. Costa and our collaborators Prof. Dr. Christoph Dieterich, Prof. Dr. Martin Zenke and Dr. Manuel Allhoff. All the results, with exception of the filter-based computational footprinting results (co-authored by Ahmad Badar), presented in this thesis were published in the following publications (in chronological order):
\\
\begin{addmargin}{0.05\textwidth}
\noindent
\textbf{Gusmao EG}, Dieterich C, Zenke M, Costa IG (2014). Detection of active transcription factor binding sites with the combination of DNase hypersensitivity and histone modifications. \textit{Bioinformatics}, 30(22):3143--3151.\\ \\
\noindent
\textbf{Gusmao EG}, Allhoff M, Zenke M, Costa IG (2016). Analysis of computational footprinting methods for DNase sequencing experiments. \textit{Nature Methods}, 13(4):303--309.\\
\end{addmargin} 

Furthermore, some parts of the following publications, co-authored by me, were used in Section~\ref{sec:case.studies} as case studies in this thesis:
\\
\begin{addmargin}{0.05\textwidth}
\noindent
Lin Q, Chauvistr\'{e} H, Costa IG, \textbf{Gusmao EG}, Mitzka S, H\"{a}enzelmann S, Baying B, Klisch T, Moriggl R, Hennuy B, Smeets H, Hoffmann K, Benes V, Ser\'{e} K, Zenke M (2015). Epigenetic and transcriptional architecture of dendritic cell development. \textit{Nucleic Acids Research}, 43(20):9680--9693.\\ \\
\noindent
Kolovos P, Georgomanolis T, Nikolic M, Koeferle A, Larkin JD, Feuerborn A, van Ijcken WF, \textbf{Gusmao EG}, Costa IG, Cook PR, Grosveld FG, Papantonis A (2016). Enhancer hijacking reveals a multimodal role of NF-$\kappa$B during the immediate-early inflammatory response. \textit{(in review)}.\\ \\

\end{addmargin} 

Finally, the following publications did not directly contribute to this thesis, but shaped a general understanding of next-generation sequencing analysis, transcription factor motif analysis and machine learning. They were published during my Ph.D. studies in RWTH University Aachen.
\\
\begin{addmargin}{0.05\textwidth}
\noindent
\textbf{Gusmao EG}, de Souto MCP (2014). Issues on Sampling Negative Examples for Predicting Prokaryotic Promoters. \textit{Proceedings of the International Joint Conference on Neural Networks (IJCNN 2014)}, 494--501. \\ \\
\noindent
Ullius A, L\"{u}scher-Firzlaff J, Costa IG, Walsemann G, Forst AH, \textbf{Gusmao EG}, Kapelle K, Kleine H, Kremmer K, Vervoorts J, L\"{u}scher B (2014). The interaction of MYC with the trithorax protein ASH2L promotes gene transcription by regulating H3K27 modification. \textit{Nucleic Acids Research}, 42(11):6901--6920. \\ \\
\noindent
H\"{a}nzelmann S, Beier F, \textbf{Gusmao EG}, Koch CM, Hummel S, Charapitsa I, Joussen S, Benes V, Br\"{u}mmendorf TH, Reid G, Costa IG, Wagner W (2015). Replicative senescence is associated with nuclear reorganization and with DNA methylation at specific transcription factor binding sites. \textit{Clinical Epigenetics}, 7(1):19. \\ \\
\noindent
Schemionek M, Herrmann O, Reher MM, Chatain N, Schubert C, Costa IG, \linebreak H\"{a}nzelmann S, \textbf{Gusmao EG}, Kintsler S, Braunschweig T, Hamilton A, Helgason GV, Copland M, Schwab A, M\"{u}ller-Tidow C, Li S, Holyoake TL, Br\"{u}mmendorf TH, \linebreak Koschmieder S (2015). MTSS1 is a critical epigenetically regulated tumor suppressor in CML. \textit{Leukemia}, 30(4):823--832. \\
\end{addmargin}

I implemented HINT and built the benchmarking dataset. All figures/tables in this thesis are authored by me and exceptions were explicitly stated in their respective captions. I performed all experiments in this study with the aid of Dr. Ivan G. Costa, Prof. Dr. Martin Zenke, Dr. Manuel Allhoff, Ahmad Badar, Dr. Sonja H\"{a}nzelmann and Joseph Kuo. All chapters of this thesis have been written by me. Ivan G. Costa provided support in all stages of this research including thesis manuscript writing. Mostly out of habit and to honor the fact that research is rarely an entirely solitary process I will, in this thesis, rely on the use of the first-person plural pronoun ``we'' in the text, as a nosism.


